---
title: "Project 1 Live Demo: Current Account Dynamics (GMD Data)"
subtitle: "Follow-along Quarto doc (Positron + Copilot + R)"
author:
  - Christopher Ball
  - Aman Desai
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: false
    code-fold: false
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

# 🎯 Goal for today

By the end of this demo, everyone will be able to:

- Open a Quarto document in Positron
- Run R code step-by-step (chunk-by-chunk)
- Use GitHub Copilot to speed up coding (without turning off your brain)
- Load, clean, and analyze the GMD Excel dataset used in Project 1
- Render this document into an HTML report you can submit/share

:::{.callout-note}
### 📁 Files you need in the SAME folder as this `.qmd`
1. `GMD_sub_1995_2019.xlsx` (the dataset used in Project 1)
2. This file: `proj1-gmd-live-demo.qmd`
:::

:::{.callout-tip}
### 💻 How to render (Positron Terminal)
```bash
# Preview while editing (auto-refreshes)
quarto preview proj1-gmd-live-demo.qmd

# Render to HTML (creates .html file)
quarto render proj1-gmd-live-demo.qmd --to html

# Render to PDF (needs LaTeX)
quarto render proj1-gmd-live-demo.qmd --to pdf
```
:::

:::{.callout-important}
### 🚀 How to follow along in class
1. Open this file in Positron
2. Click on any code chunk
3. Press `Ctrl+Enter` (Windows/Linux) or `Cmd+Enter` (Mac) to run that chunk
4. Watch the output appear below the chunk
5. Move to next chunk and repeat
:::

---

# Step 1 — Install/Load Packages

Run this chunk FIRST. If you get errors, you may need to install packages first.

```{r}
#| label: setup-packages

# Load required packages
library(tidyverse)   # for data manipulation and plotting
library(readxl)      # for reading Excel files
library(writexl)     # for writing Excel files
library(glue)        # for string formatting

```

:::{.callout-tip}
### 🤖 Copilot prompt to try
Type this comment in a code chunk and press `Tab` or `Ctrl+I`:

Copilot will suggest code to auto-install missing packages!
:::

# Step 2 — Student Inputs (EDIT THESE!)

**👉 Students: Change these three variables for YOUR project**

```{r}
#| label: student-inputs

# CHANGE THESE VALUES FOR YOUR PROJECT
my_country <- "Thailand"    # Pick from list below
start_date <- 2000          # Earliest: 1995
end_date <- 2019            # Latest: 2019 (2018 for Ireland/Mexico)

data_file <- "GMD_sub_1995_2019.xlsx"
```

:::{.callout-note}
### 📋 20 Ideal Country Options (Pick ONE!)

"Argentina", "Australia", "Brazil", "Chile", "Colombia", "Costa Rica", "Hungary", "Iceland", "Indonesia", "Ireland" (2018), "Malaysia", "Mexico" (2018), "New Zealand", "Peru", "Philippines", "Poland", "South Africa", "South Korea", "Sri Lanka", "Thailand"

**Demo countries:** "Canada", "United Kingdom", "United States"
:::

```{r}
#| label: check-file-exists

# Check if data file exists
if (!file.exists(data_file)) {
  stop(glue(
    "❌ Cannot find `{data_file}` in the current folder.\n",
    "Fix: Put the Excel file in the same folder as this .qmd, then re-run."
  ))
} else {
  glue("✅ Found data file: {data_file}")
}
```

---

# Step 3 — Read and Filter Data

This loads the Excel file and filters it to your chosen country and years.

```{r}
#| label: load-data

Proj1 <- read_excel(data_file) %>%
  filter(year >= start_date, year <= end_date) %>%
  filter(countryname == my_country)

# Show first few rows
head(Proj1)
```

```{r}
#| label: data-summary

# Quick data check
Proj1 %>%
  summarise(
    country = first(countryname),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE),
    n_rows = n(),
    n_columns = ncol(.)
  ) %>%
  knitr::kable(caption = "Data Summary")
```

:::{.callout-warning}
### ⚠️ If you get 0 rows
That means:
- Country name doesn't match exactly (check spelling/capitalization)
- Year range is outside available data
- Data file is wrong/corrupted
:::

---

# PART ONE: Basic Macro and Current Account Dynamics

---

## 1A. Current Account vs Trade Balance (Time Series)

**Theory:** Current Account (CA) = Trade Balance (TB) + Income Balance (IB)

Let's plot CA and TB together to see how closely they move.

```{r}
#| label: plot-ca-tb-time
#| fig-cap: "Current Account and Trade Balance over time"

ggplot(Proj1, aes(x = year)) +
  geom_line(aes(y = CA, color = "Current Account"), linewidth = 1) +
  geom_line(aes(y = trade_balance, color = "Trade Balance"), linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(
    title = paste(my_country, ": Current Account and Trade Balance"),
    subtitle = paste(start_date, "–", end_date),
    x = NULL,
    y = "Millions (local currency)",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

:::{.callout-tip}
### 📊 What to look for:
- Do CA and TB move together?
- Are they close in value?
- When do they diverge?
- Are both positive, both negative, or mixed?
:::

---

## 1B. Current Account vs Trade Balance (Scatter Plot)

This shows the relationship between CA and TB more clearly.

```{r}
#| label: plot-ca-tb-scatter
#| fig-cap: "Scatter plot of Current Account vs Trade Balance"

ggplot(Proj1, aes(x = CA, y = trade_balance)) +
  geom_point(size = 3, alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(
    title = paste(my_country, ": Current Account vs Trade Balance"),
    subtitle = "Red dashed line: perfect 1:1 relationship",
    x = "Current Account",
    y = "Trade Balance"
  ) +
  theme_minimal(base_size = 12)
```

:::{.callout-note}
### 🔍 Interpretation guide:
- **Points on red line:** CA = TB (no income balance)
- **Points above red line:** TB > CA (negative income balance)
- **Points below red line:** CA > TB (positive income balance)
:::

---

## 1C. Correlation Between CA and TB

Calculate the correlation coefficient to quantify the relationship.

```{r}
#| label: correlation-ca-tb

cor_CA_TB <- cor(
  x = Proj1$CA,
  y = Proj1$trade_balance,
  method = "pearson",
  use = "complete.obs"
)

glue("Correlation (CA, TB) = {round(cor_CA_TB, 4)}")
```

:::{.callout-tip}
### 🤖 Copilot prompt to try:
```r
# Explain why the correlation between CA and TB is close to but not exactly 1
```
:::

---

## 1D. Calculate Income Balance

**Income Balance = CA − TB**

This represents net income from foreign assets (interest, dividends, etc.)

```{r}
#| label: calculate-income-balance

Proj1 <- Proj1 %>%
  mutate(income_balance = CA - trade_balance)

# Show summary statistics
Proj1 %>%
  summarise(
    mean_IB = mean(income_balance, na.rm = TRUE),
    median_IB = median(income_balance, na.rm = TRUE),
    min_IB = min(income_balance, na.rm = TRUE),
    max_IB = max(income_balance, na.rm = TRUE)
  ) %>%
  knitr::kable(digits = 2, caption = "Income Balance Summary Statistics")
```

---

## 1E. Plot All Three: CA, TB, and Income Balance

```{r}
#| label: plot-ca-tb-ib
#| fig-cap: "Current Account decomposition"

ggplot(Proj1, aes(x = year)) +
  geom_line(aes(y = CA, color = "Current Account"), linewidth = 1) +
  geom_line(aes(y = trade_balance, color = "Trade Balance"), linewidth = 1) +
  geom_line(aes(y = income_balance, color = "Income Balance"), linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(
    title = paste(my_country, ": CA, TB, and Income Balance"),
    subtitle = paste(start_date, "–", end_date),
    x = NULL,
    y = "Millions (local currency)",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

```

:::{.callout-important}
### 💡 Key Question for Class Discussion:
**Q:** What explains the difference between CA and TB?  
**A:** Income Balance (r × B), which represents international asset earnings/payments
:::

---



# PART TWO: Absorption Approach (GDP vs Domestic Absorption)

---

## 2A. Calculate Domestic Absorption

**Domestic Absorption = C + I + G**

This is total domestic spending (consumption + investment + government).

```{r}
#| label: calculate-absorption

Proj1 <- Proj1 %>%
  mutate(dom_abs = cons + inv + govexp)

# Quick check
Proj1 %>%
  select(year, nGDP, dom_abs, CA) %>%
  head() %>%
  knitr::kable(digits = 0, caption = "GDP, Absorption, and CA (first 6 years)")
```

:::{.callout-note}
### 📚 Theory reminder:
**CA = GDP − (C + I + G)**

Or equivalently: **CA = GDP − Domestic Absorption**

- When absorption > GDP → CA is negative (borrowing)
- When GDP > absorption → CA is positive (lending)
:::

---

## 2B. Plot GDP, Absorption, and CA (Faceted)

```{r}
#| label: plot-gdp-absorption-ca
#| fig-cap: "Income, Domestic Absorption, and Current Account"
#| fig-height: 8

Proj1_long_abs <- Proj1 %>%
  select(year, nGDP, dom_abs, CA) %>%
  pivot_longer(
    cols = c(nGDP, dom_abs, CA),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    panel = case_when(
      variable %in% c("nGDP", "dom_abs") ~ "Income and Domestic Absorption",
      variable == "CA" ~ "Current Account"
    ),
    variable = recode(
      variable,
      nGDP = "Nominal GDP",
      dom_abs = "Domestic Absorption",
      CA = "Current Account"
    )
  )

ggplot(Proj1_long_abs, aes(x = year, y = value, color = variable)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~ panel, ncol = 1, scales = "free_y") +
  labs(
    title = paste(my_country, ": Income, Absorption, and the Current Account"),
    subtitle = paste(start_date, "–", end_date),
    x = NULL,
    y = NULL,
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

:::{.callout-tip}
### 🤖 Copilot prompt:
```r
# Write a paragraph explaining what it means when domestic absorption exceeds GDP
```
:::

---

# PART THREE: Savings-Investment Approach

---

## 3A. Calculate National Savings

**National Savings = GDP − C − G**

(We subtract consumption and government spending from GDP)

```{r}
#| label: calculate-savings

Proj1 <- Proj1 %>%
  mutate(savings = nGDP - (cons + govexp))

# Summary stats
Proj1 %>%
  summarise(
    mean_savings = mean(savings, na.rm = TRUE),
    mean_investment = mean(inv, na.rm = TRUE),
    mean_CA = mean(CA, na.rm = TRUE)
  ) %>%
  knitr::kable(digits = 0, caption = "Average Savings, Investment, and CA")
```

:::{.callout-note}
### 📚 Theory reminder:
**CA = S − I**

- When S > I → CA positive (country is lending/saving abroad)
- When I > S → CA negative (country is borrowing to invest)
:::

---

## 3B. Plot Savings, Investment, and CA

```{r}
#| label: plot-savings-investment-ca
#| fig-cap: "Savings, Investment, and Current Account"
#| fig-height: 8

Proj1_sav_inv_long <- Proj1 %>%
  select(year, savings, inv, CA) %>%
  pivot_longer(
    cols = c(savings, inv, CA),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    panel = if_else(variable == "CA", "Current Account", "Savings and Investment"),
    variable = recode(
      variable,
      savings = "Savings",
      inv = "Investment",
      CA = "Current Account (CA)"
    )
  )

ggplot(Proj1_sav_inv_long, aes(x = year, y = value, color = variable)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~ panel, ncol = 1, scales = "free_y") +
  labs(
    title = paste(my_country, ": Savings, Investment, and the Current Account"),
    subtitle = paste(start_date, "–", end_date),
    x = NULL,
    y = NULL,
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

# PART FOUR: Business Cycle Correlations

---

## 4A. Correlation Matrix

Let's see how GDP, absorption, and trade balance move together.

```{r}
#| label: correlation-matrix

cor_matrix <- Proj1 %>%
  select(nGDP, dom_abs, trade_balance, CA) %>%
  cor(use = "complete.obs")

cor_matrix %>%
  knitr::kable(digits = 3, caption = "Correlation Matrix")
```

---

## 4B. GDP vs Trade Balance

```{r}
#| label: correlations-gdp

cor_gdp_tb <- cor(Proj1$nGDP, Proj1$trade_balance, use = "complete.obs")
cor_gdp_ca <- cor(Proj1$nGDP, Proj1$CA, use = "complete.obs")

glue("Correlation (GDP, TB) = {round(cor_gdp_tb, 4)}")
glue("Correlation (GDP, CA) = {round(cor_gdp_ca, 4)}")
```

:::{.callout-important}
### 🔍 Interpretation:
- **Negative correlation:** Trade balance/CA moves opposite to GDP (counter-cyclical)
- **Positive correlation:** Trade balance/CA moves with GDP (pro-cyclical)
:::

---

## 4C. Scatter: GDP vs Trade Balance

```{r}
#| label: plot-gdp-tb-scatter
#| fig-cap: "GDP vs Trade Balance relationship"

ggplot(Proj1, aes(x = nGDP, y = trade_balance)) +
  geom_point(size = 3, alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = paste(my_country, ": Nominal GDP vs Trade Balance"),
    subtitle = paste(start_date, "–", end_date),
    x = "Nominal GDP",
    y = "Trade Balance"
  ) +
  theme_minimal(base_size = 12)
```

---

## 4D. Scatter: Domestic Absorption vs Trade Balance

```{r}
#| label: plot-absorption-tb-scatter
#| fig-cap: "Domestic Absorption vs Trade Balance"

cor_abs_tb <- cor(Proj1$dom_abs, Proj1$trade_balance, use = "complete.obs")

ggplot(Proj1, aes(x = dom_abs, y = trade_balance)) +
  geom_point(size = 3, alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = paste(my_country, ": Domestic Absorption vs Trade Balance"),
    subtitle = glue("Correlation = {round(cor_abs_tb, 3)}"),
    x = "Domestic Absorption",
    y = "Trade Balance"
  ) +
  theme_minimal(base_size = 12)
```

---

## 4E. Time Series: GDP, Absorption, and TB

```{r}
#| label: plot-gdp-abs-tb-time
#| fig-cap: "GDP, Absorption, and Trade Balance over time"
#| fig-height: 8

Proj1_long_cyc <- Proj1 %>%
  select(year, nGDP, dom_abs, trade_balance) %>%
  pivot_longer(
    cols = c(nGDP, dom_abs, trade_balance),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    panel = case_when(
      variable %in% c("nGDP", "dom_abs") ~ "Income and Domestic Absorption",
      variable == "trade_balance" ~ "Trade Balance"
    ),
    variable = recode(
      variable,
      nGDP = "Nominal GDP",
      dom_abs = "Domestic Absorption",
      trade_balance = "Trade Balance"
    )
  )

ggplot(Proj1_long_cyc, aes(x = year, y = value, color = variable)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~ panel, ncol = 1, scales = "free_y") +
  labs(
    title = paste(my_country, ": Income, Absorption, and the Trade Balance"),
    subtitle = paste(start_date, "–", end_date),
    x = NULL,
    y = NULL,
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

# PART FIVE: Government Balance

---

## 5A. Government Revenues, Expenditures, and Deficit

```{r}
#| label: plot-government-budget
#| fig-cap: "Government fiscal position"
#| fig-height: 8

Proj1_gov_long <- Proj1 %>%
  select(year, govrev, govexp, govdef) %>%
  pivot_longer(
    cols = c(govrev, govexp, govdef),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    panel = if_else(variable == "govdef",
                    "Government Deficit",
                    "Government Revenues and Expenditures"),
    variable = recode(
      variable,
      govrev = "Government Revenue",
      govexp = "Government Expenditure",
      govdef = "Government Deficit"
    )
  )

ggplot(Proj1_gov_long, aes(x = year, y = value, color = variable)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~ panel, ncol = 1, scales = "free_y") +
  labs(
    title = paste(my_country, ": Government Revenues, Expenditures, and Deficit"),
    subtitle = paste(start_date, "–", end_date),
    x = NULL,
    y = NULL,
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 5B. Government Deficit and Current Account

Are "twin deficits" present? (Government deficit and CA deficit moving together)

```{r}
#| label: plot-gov-ca
#| fig-cap: "Government balance vs Current Account"
#| fig-height: 8

Proj1_govdef_ca_long <- Proj1 %>%
  select(year, govdef, CA) %>%
  pivot_longer(
    cols = c(govdef, CA),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    variable = recode(
      variable,
      govdef = "Government Deficit",
      CA = "Current Account"
    )
  )

ggplot(Proj1_govdef_ca_long, aes(x = year, y = value, color = variable)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~ variable, ncol = 1, scales = "free_y") +
  labs(
    title = paste(my_country, ": Government Balance and Current Account"),
    subtitle = paste(start_date, "–", end_date),
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

```{r}
#| label: correlation-gov-ca

cor_govdef_ca <- cor(Proj1$govdef, Proj1$CA, use = "complete.obs")
glue("Correlation (Gov Deficit, CA) = {round(cor_govdef_ca, 4)}")
```

---

## 5C. As % of GDP

```{r}
#| label: plot-gov-ca-pct-gdp
#| fig-cap: "Government deficit and CA as % of GDP"
#| fig-height: 8

Proj1_govdef_ca_gdp_long <- Proj1 %>%
  select(year, govdef_GDP, CA_GDP) %>%
  pivot_longer(
    cols = c(govdef_GDP, CA_GDP),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    variable = recode(
      variable,
      govdef_GDP = "Government Deficit (% GDP)",
      CA_GDP = "Current Account (% GDP)"
    )
  )

ggplot(Proj1_govdef_ca_gdp_long, aes(x = year, y = value, color = variable)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~ variable, ncol = 1, scales = "free_y") +
  labs(
    title = paste(my_country, ": Government Balance (% GDP) and Current Account (% GDP)"),
    subtitle = paste(start_date, "–", end_date),
    x = NULL,
    y = "Percent of GDP"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

---

# PART SIX: Export Cleaned Data

---

## 6. Save Your Work

Export the cleaned dataset with all calculated variables.

```{r}
#| label: export-data

# Create filename
clean_file <- paste0(
  "Proj1_clean_",
  gsub(" ", "_", my_country),
  "_",
  start_date,
  "_",
  end_date,
  ".xlsx"
)

# Write to Excel
write_xlsx(Proj1, clean_file)

glue("✅ Exported cleaned data to: {clean_file}")
```

---

# Summary and Key Takeaways

---

## What We Learned Today

1. ✅ How to load and filter Excel data in R
2. ✅ Current Account = Trade Balance + Income Balance
3. ✅ CA = GDP − (C + I + G) (absorption approach)
4. ✅ CA = S − I (savings-investment approach)
5. ✅ How to create professional time-series and scatter plots
6. ✅ How to calculate correlations and interpret them
7. ✅ How to use Copilot to speed up coding
8. ✅ How to render a Quarto document to HTML/PDF

---

## Student Checklist (What to Submit)

:::{.callout-important}
### 📋 Before you submit:

1. ✅ Change `my_country`, `start_date`, `end_date` to YOUR chosen values
2. ✅ Re-run ALL code chunks from top to bottom
3. ✅ Render to HTML: `quarto render proj1-gmd-live-demo.qmd --to html`
4. ✅ Check that all plots show YOUR country's data
5. ✅ Submit:
   - Your rendered `.html` file
   - Your edited `.qmd` file
   - Your cleaned `.xlsx` export (optional)
   - Your GitHub repo link (if applicable)
:::

---

## Copilot Prompts You Can Try

:::{.callout-tip}
### 🤖 Useful prompts for your project:

**For understanding:**
- "Explain what the current account measures in plain English"
- "Why would a country have a negative current account?"
- "What's the difference between trade balance and current account?"

**For coding:**
- "Add a column that calculates CA as percent of GDP"
- "Create a scatter plot with a regression line"
- "Write code to export only specific columns to Excel"

**For writing:**
- "Write a paragraph explaining why this country's CA is negative/positive"
- "Summarize the relationship between government deficit and CA in 3 sentences"
- "What does a correlation of X mean in this context?"
:::

---

## Resources

- **Quarto Documentation:** https://quarto.org/docs/
- **GitHub Copilot for Students:** https://education.github.com/pack
- **Positron IDE:** https://positron.posit.co
- **ggplot2 Reference:** https://ggplot2.tidyverse.org
- **Your instructor:** YOUR-EMAIL

---

## Questions?

:::{.callout-note}
### 💬 After class:
- Email me with questions: YOUR-EMAIL
- Office hours: [YOUR DAYS/TIMES]
- Post in course discussion board
- Work with classmates!
:::

---

**🎉 Congratulations! You've completed the live demo!**

Now try changing the country, re-running all chunks, and rendering your own version!

```{r}
#| echo: false

cat("✅ Demo complete! Now customize and render your own version.\n")
```
